# Inflation Developement Analysis

## Context

With the present document, I'll explore the data available, gather insights and inspect patterns in order to responde to the following questions:

> How can we statistically characterize the behavior of inflation across countries, and has it significantly changed since 2000?

More specifically, we would like to explore the following sub-objectives:

1.  Are inflation distribution flat or fat-tailed?
2.  Is inflation stationary or is there seasonal patterns?
3.  How does crisis influence inflation markers?
4.  Which indicators could be used to predict inflation rates?

Before diving into the exploration, it is necessary to merge both available datasets into a single one, as well as do a simple cleaning.

```{r}
library(readr)
library(dplyr)
```


## Preprocessing

As mentioned before, the first step is the merging of both datasets.
As demonstrated below, there are multiple condidates for merging keys:

* 'year': containing unique year value from 2000 to 2024
* 'country': containing 265 countries around the world name in the english locale.
* 'iso2c' and 'iso3c', which are country identifiers.

As the 'comp' data is only complementary to 'main', a right join on main is sensible. In order to avoid duplicated columns, I'll join those 4 possible identifiers.

```{r}
main <- readRDS("data/raw/main.rds")
comp <- readRDS("data/raw/comp.rds")
```

```{r}
print(paste0("Length of the 'year' column in `main`: ", length(main$year)))
print(paste0("Length of the 'year' column in `comp`: ", length(comp$year)))

print(paste0("'year' in main ranges from ", min(main$year), "to ", max(main$year)))
print(paste0("'year' in main ranges from ", min(comp$year), "to ", max(comp$year)))
```

```{r}
print(paste0("Amount of recorded unique countries in `main`: ", length(unique(main$country)), " for ", length(main$country), " total observations"))
print(paste0("Amount of recorded unique countries in `comp`: ", length(unique(comp$country)), " for ", length(comp$country), " total observations"))

print(paste0("Is Switzerland named 'Switzerland' (EN)? ", ("Switzerland" %in% unique(main$country)) & ("Switzerland" %in% unique(comp$country))))
```

```{r}
join <- right_join(main, comp, by=c("year", "country", "iso2c", "iso3c"))
```


# Sources

Our source of datasets and indicators: [World Bank Group](https://data.worldbank.org/indicator)
